Calibrating MEM seeding heuristics

### 2018-04-06 MEM.tex
Clearing the mess.

\newcommand{\smA}{\scriptstyle{A}}
\newcommand{\smC}{\scriptstyle{C}}
\newcommand{\smD}{\scriptstyle{D}}
\newcommand{\smE}{\scriptstyle{E}}
\newcommand{\smG}{\scriptstyle{G}}
\newcommand{\smI}{\scriptstyle{I}}
\newcommand{\smS}{\scriptstyle{S}}
\newcommand{\smT}{\scriptstyle{T}}
\newcommand{\smDELz}{\scriptstyle{\Delta_0}}
\newcommand{\smDELs}{\scriptstyle{\Delta_*}}

@@

@@

\begin{equation*}
M(z) =
\begin{blockarray}{ccccccc}
   & \scriptstyle{\blacksquare} & \scriptstyle{\times} 
   & \scriptstyle{(!)_1} & \scriptstyle{(!)_2}
   & \ldots & \scriptstyle{(!)_{\gamma-1}} \\
\begin{block}{c[cccccc]}
\scriptstyle{\blacksquare} & \tilde{A}_\infty(z) & \tilde{B}_\infty(z) &
    r_1(z) & r_2(z) & \ldots & r_{\gamma-1}(z) \\
\scriptstyle{\times} & \tilde{A}_\infty(z) & \tilde{B}_\gamma(z) &
    r_1(z) & r_2(z) & \ldots & r_{\gamma-1}(z) \\
\scriptstyle{(!)_1} & A_{\gamma-1}(z) & B_{\gamma-1}(z) & 0 & 0 &
    \ldots & 0 \\
\scriptstyle{(!)_2} & A_{\gamma-2}(z) & B_{\gamma-2}(z) & 0 & 0 &
    \ldots & 0  \\
\vdots & \vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
\scriptstyle{(!)_{\gamma-1}} & A_1(z) & B_1(z) & 0 & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

\begin{eqnarray*}
A_n(z) &=& p (1-(1-\kappa/3)^N) \sum_{m=0}^{n-1} q^mz^{m+1}, \\
\tilde{A}_n(z) &=& p (1-(1-\kappa/3)^N)
    \sum_{m=0}^{n-1} C_m q^mz^{m+1}, \\
B_n(z) &=& p(1-\kappa/3)^N \sum_{m=0}^{n-1} q^mz^{m+1}, \\
\tilde{B}_n(z) &=& p(1-\kappa/3)^N \sum_{m=0}^{n-1} C_m q^mz^{m+1}, \\
r_n(z) &=& (C_{n-1}-C_n) (qz)^n, \\
F_n(z) &=& \sum_{m=0}^n C_m (qz)^m.
\end{eqnarray*}

Define $C_m = \big( 1- (1-(1-\kappa)^m)^N \big)$ as the probability that
during $m$ nucleotides, not all duplicates have collapsed.

>>>>

\newcommand{\DN}{\scriptstyle{\Downarrow}}
\newcommand{\dn}{\scriptstyle{\downarrow}}
\newcommand{\up}[1]{\scriptstyle{(\uparrow)_{#1}}}
\newcommand{\eq}[1]{\scriptstyle{(\sim)_{#1}}}

\newcommand{\upg}{\scriptstyle{\uparrow}}
\newcommand{\eqg}{\scriptstyle{\sim}}

@@

\section{Some definitions etc.}

@@

We assume that the target sequence has $N$ duplicates that are potential
false positives for the seeding process. We further assume that
duplication was instantaenous and that all $N+1$ sequences diverge
independently from each other at a constant rate. In other words, we
ignore the complications due to the genealogy of the duplication events
and we simply assume that at each position, the target and a duplicate are
identical with probability $1-\mu$. In case they are not, we assume that
the duplicate has any of the remaining three nucleotides with equal
probability.

Each of the $N+1$ sequences of interest corresponds to a Bernoulli process
where `success' stands for a match between the sequence and the read, and
`failure' stands for a mismatch. To describe those processes, we will use
the `match function' $\varphi(n)$ that gives the length of the success
streak at position $n$. The processes are indexed such that $\varphi_0$
describes the matches of the target sequence, whereas $\varphi_j$ $(1 \leq
j \leq N)$ describes the matches of the duplicates. For instance,
$\varphi_0(5) = 5$ and $\varphi_1(5) = 3$ in the example of
Fig.~\ref{fig:sketch_MEM}.

For simplicity, we will refer to the Bernoulli processes as `threads', and
we will distinguish the `target thread' from the `duplicate threads'. At
any position $n$ of the read, a duplicate thread $j$ $(1 \leq j \leq N)$
is said to be `strictly masking' if $\varphi_j(n) > \varphi_0(n)$,
`potentially masking' if $\varphi_j(n) \geq \varphi_0(n)$ and
`non-masking' if $\varphi_j(n) < \varphi_0(n)$. There can be no MEM seed
at positions where there is a stricly masking thread. At positions where
there is a potentially masking thread any MEM seed must be a shared MEM
seed.

With these definitions, the decision tree for choosing the next symbol to
insert in a read can be summarized as in Fig.~\ref{fig:decision_tree}.
This sketch will be useful as a reference to compute the probabilities of
the blocks and their weighted generating functions.

\begin{figure}[h]
\centering
\includegraphics[scale=.7]{decision_tree.pdf}
\caption{\textbf{Decision tree}. 
Some legend.}
\label{fig:decision_tree}
\end{figure}

\begin{equation*}
\begin{blockarray}{ccccccccc}
   & \dn & \DN &\eq{1} & \ldots & \eq{\gamma-1} &
    \up{1} & \ldots & \up{\gamma-1} \\
\begin{block}{c[cccccccc]}
\dn & * & * & * & \ldots & * & * & \ldots & * \\
\DN & \tilde{A}_\infty(z) & \tilde{B}_\gamma(z) & 0 & \ldots & 0 &
    r_1(z) & \ldots & r_{\gamma-1}(z) \\
\eq{1} & * & * & 0 & \ldots & 0 & 0 & \ldots & G_{1,\gamma-1}(z) \\
\eq{2} & * & * & 0 & \ldots & 0 & 0 & \ldots & G_{2,\gamma-1}(z) \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\eq{\gamma-1} & * & * & 0 & \ldots & 0 & 0 & \ldots & 0 \\
\up{1} & A_{\gamma-1}(z) & B_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\up{2} & A_{\gamma-2}(z) & B_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0  \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\up{\gamma-1} & A_1(z) & B_1(z) & 0 & \ldots & 0 & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

\begin{definition}
The main thread is dominant at nucleotide $n$ if $\varphi_0(n) >
\varphi_i(n)$ for all $i$ $(1 \leq i \leq N)$. The main thread is
co-dominant if inequality is not strict and there exists $i$ $(1\leq i
\leq N)$ such that $\varphi_0(n) = \varphi_i(n)$.
\end{definition}

\subsection{Head vector}

At the beginning of the read, the match lengths of all the threads are
$0$. This is exactly the same as on a $\Downarrow$ symbol. We can thus
consider that the read starts in the $\Downarrow$ state, which is
equivalent to adding an edge labelled with the empty object $\varepsilon$
between the head vertex and the $\Downarrow$ vertex. The head vector is
thus
\begin{equation}
H(z) = (1,0, \ldots, 0).
\end{equation}

\subsection{Transitions from $\uparrow$}
\label{sec:trans_from_up}

After a $\uparrow$ symbol, the main thread becomes dominant. It will
remain so until the first read error, which can be a mismatch for all the
threads ($\Downarrow$ symbol), or for only some of them ($\downarrow$
symbol).

Since the read has no MEM seed, the $\Downarrow$ symbol or the
$\downarrow$ symbol must come before $\varphi_0(n) \geq \gamma$, which
imposes a limit on the number of $\square$ symbols that can follow a
$\square$ symbol. If nucleotide at position $n$ is the $\uparrow$ symbol
and $\varphi_0(n) = j$ $(1 \leq j \leq \gamma-1)$, the state of the read
is represented as $(\uparrow)_j$, meaning that the maximum number of
$\square$ symbols that can follow is $\gamma-1-j$.

Whether the nucleotides match the alternative threads is irrelevant until
the end of the block. The $\square$ symbols only mean that the nucleotides
are not error so their weighted generating function is $qz$. The
$\Downarrow$ symbol indicates an error -- probability $p$ -- that is also
a mismatch for the $N$ alternative threads -- probability $(1-\mu/3)^N$ --
so its weighted generating function is $pz(1-\mu/3)^N$. Likewise, the
weighted generating function of the $\downarrow$ symbol is seen to be
$pz(1-(1-\mu/3)^N)$.

The weighted generating function of the transition from $(\uparrow)_j$ to
$\Downarrow$ is
\begin{equation}
B_{\gamma-j}(z) = pz (1-\mu/3)^N \sum_{n=0}^{\gamma-j-1} (qz)^n.
\end{equation}

And the weighted generating function of the transition from
$(\uparrow)_j$ to $\downarrow$ is
\begin{equation}
A_{\gamma-j}(z) = pz (1-(1-\mu/3)^N) \sum_{n=0}^{\gamma-j-1} (qz)^n.
\end{equation}

\subsection{Transitions from $\Downarrow$}
\label{sec:trans_from_Down}

At a $\Downarrow$ symbol, all the threads have match length $0$ and the
read is `reset' to its initial state. The main thread is co-dominant with
all the alternative threads. The next terminator symbol depends on what
happens first. If all the alternative threads fail before the main thread,
the latter becomes dominant ($\uparrow$ symbol). If not all alternative
threads have failed when the main thread does, we need to distinguish two
cases: if all threads fail together with the main thread, the read is
reset ($\Downarrow$ symbol), otherwise the main read becomes dominated
($\downarrow$ symbol).

If the next terminator is the $\uparrow$ symbol, we need to distinguish
the cases based on the value of $\varphi_0$ (see
section~\ref{sec:trans_from_up}), which is just the position of
the $\uparrow$ symbol after the $\Downarrow$ symbol in this case. A
transition from $\Downarrow$ to $(\uparrow)_j$ consists of a succession of
$j$ correct nucleotides -- each with probability $q$. But we must also
make sure that no other $\uparrow$ symbol is inserted before position $j$.
We thus introduce $\xi_j$, the probability that at least one of the $N$
alternative threads survives at position $j$ after the $\Downarrow$ symbol. At
any correct nucleotide, the probability of failure of an alternative thread is
$\mu$, so
\begin{equation}
\xi_j = 1-(1-(1-\mu)^j)^N.
\end{equation}

The probablity of transition from $\Downarrow$ to $(\uparrow)_j$ is thus
$(\xi_{j-1}-\xi_j)q^j$ and the associated weighted generating function is
\begin{equation}
r_j(z) = (\xi_{j-1}-\xi_j)(qz)^j.
\end{equation}

If the next terminator is $\downarrow$, there is no limit on the number of
$\square$ symbols that can be inserted. Indeed, the main thread will not
be part of the MEM in this case. As mentioned in
section~\ref{sec:trans_from_up}, the weighted generating function of the
$\downarrow$ symbol is $pz(1-(1-\mu/3)^N)$. The probability that it first
occurs at position $n+1$ after the $\Downarrow$ symbol is the probability
that the first $n$ nucleotides are correct -- each with probability $q$
-- and that at least one alternative thread survives -- probability
$\xi_n$. The weighted generating function of the transition from
$\Downarrow$ to $\downarrow$ is thus
\begin{equation}
\tilde{A}_\infty(z) = pz(1-(1-\mu/3)^N)
  \sum_{n=0}^\infty \xi_n \cdot (qz)^n.
\end{equation}

Finally, if the next terminator is $\Downarrow$, there is a limit on the
number of $\square$ symbols. Indeed, if there are $\gamma$ or more
nucleotides between the two $\Downarrow$ symbols, they constitute a shared
MEM seed. As mentioned in section~\ref{sec:trans_from_up}, the weighted
generating function of the $\Downarrow$ symbol is $pz(1-\mu/3)^N$. By the
same rationale as above, the weighted generating function of the
transition from $\Downarrow$ to $\Downarrow$ is
\begin{equation}
\tilde{B}_{\gamma}(z) = pz(1-\mu/3)^N
  \sum_{n=0}^{\gamma-1} \xi_n \cdot (qz)^n.
\end{equation}

\subsection{Transitions from $\sim$}

At a $\sim$ symbol, the main thread becomes co-dominant, together with $1$
to $N-1$ alternative threads. As in section~\ref{sec:trans_from_Down}, the
next terminator symbol depends on whether all the alternative co-dominant
threads fail before the main thread. If this is the case, the main thread
will become dominant ($\uparrow$ symbol), otherwise we again need to
distinguish the case that the read is reset ($\Downarrow$ symbol) from the
case that the main thread becomes dominated ($\downarrow$ symbol).

An important difficulty is that the $\sim$ symbol does not inform us about
the number of surviving threads (except when $N = 1$). So we need to
marginalize the probabilities of interest over this number. The first
thing to note is that a $\sim$ terminator can only follow a $\downarrow$
terminator. Indeed, after a $\Downarrow$ symbol the threads are already
co-dominant, and after a $\uparrow$ symbol the alternative threads cannot
`catch up' with the main thread until a read error occurs, \textit{i.e.}
until a $\Downarrow$ symbol or a $\downarrow$ symbol is inserted. The
number of surviving threads depends on the position of the $\sim$ symbol
relative to the preceding $\downarrow$ symbol, so as in
section~\ref{sec:trans_from_up} we need to distinguish states based on the
value of $\varphi_0$ at the position of the $\sim$ symbol.

If nucleotide at position $n$ is $\sim$ and $\varphi_0(n) = j$ $(1 \leq j
\leq \gamma-1)$, the state of the read is represented as $(\sim)_j$.

\subsection{Transitions from $\downarrow$}

Like... more complicated.

By conditioning on how many duplicate threads fail at the $\dn$ symbol, we
can compute the probability that a $\eqg$ symbol occurs at position $k$
after an error as $q^kE_k / (1-(1-\mu/3)^N)$, where

\begin{equation*}
\begin{split}
E_k = \sum_{m=1}^{N-1} {N \choose m} (1-\mu/3)^m (\mu/3)^{N-m} \times
\Big( 1 - (1-(1-\mu)^k)^m \Big) \times \\
\Big( (1-(1-\mu)^k)^{N-m} - (1-(1-\mu)^{k-1})^{N-m} \Big).
\end{split}
\end{equation*}

\begin{equation*}
\begin{split}
E_k = [1-\mu/3(1-\mu)^k]^N - [1-\mu/3(1-\mu)^{k-1}]^N
-[1-(1-\mu)^k]^N + \\
[1-(1-\mu+\mu^2/3)(1-\mu)^{k-1}]^N.
\end{split}
\end{equation*}

The calculation above says that $m$ threads fail at the $\dn$ symbol, that
they do not all fail in the following $k$ nucleotides, and that the last
survivor of the $\dn$ symbol fails at nucleotide $k$. Note that $E_k=0$
for $N=0$ and $N=1$, as expected. To find the probability, we need to
multiply by $q^k$ to account for the fact that there is no error in $k$
nucleotides, and we need to divide by $(1-(1-\mu/3)^N)$ to account for the
fact that not all threads can fail at symbol $\dn$ (otherwise it would be
symbol $\DN$). The sum does not include the case $m=0$ because in this
case, none of the duplicate threads can have the same length as the main
thread, so there cannot be a $\eqg$ symbol.

\begin{eqnarray*}
\sum_{m=n}^N {N \choose m} {m \choose n} x^m y^{N-m}
&=& x^n \sum_{r=0}^{N-n} {N \choose n+r} {n+r \choose n}a^r y^{N-n-r} \\
&=& {N \choose n} x^n \sum_{r=0}^{N-n} {N-n \choose r}x^r y^{N-n-r} \\
&=& {N \choose n} x^n (x+y)^{N-n}.
\end{eqnarray*}

\begin{equation*}
\sum_{m=n}^{N-1} {N \choose m} {m \choose n} x^m y^{N-m} =
{N \choose n} x^n \Big[ (x+y)^{N-n} - x^{N-n} \Big].
\end{equation*}

The probability of symbol $\eqg$ with $n$ surviving threads after
$\scriptstyle{\blacksquare}$ in $k$ steps is
\begin{equation*}
\begin{split}
\sum_{m=n}^{N-1} {N \choose m} (1-\mu/3)^m (\mu/3)^{N-m} \times
{m \choose n} (1-\mu)^{kn} \big(1-(1-\mu)^k\big)^{m-n} \times \\
\Big( (1-(1-\mu)^k)^{N-m} - (1-(1-\mu)^{k-1})^{N-m} \Big).
\end{split}
\end{equation*}

We obtain a first term equal to
\begin{equation*}
(1-\mu)^{kn} (1-(1-\mu)^k)^{N-n}
\sum_{m=n}^{N-1} {N \choose m} {m \choose n} (1-\mu/3)^m (\mu/3)^{N-m}.
\end{equation*}

Using the formula above with $x=1-\mu/3$ and $y=\mu/3$, we see that this
first term is equal to
\begin{equation*}
{N \choose n} (1-\mu)^{kn} (1-\mu/3)^n \bigg( \big[ 1-(1-\mu)^k \big]^{N-n}
-\big[ (1-(1-\mu)^k)(1-\mu/3)\big]^{N-n} \bigg)
\end{equation*}

The second term is
\begin{equation*}
\begin{split}
\left(\frac{(1-\mu)^k}{1-(1-\mu)^k}\right)^n \cdot
\sum_{m=n}^{N-1} {N \choose m} {m \choose n} (1-\mu/3)^m (\mu/3)^{N-m}
\times \\
(1-(1-\mu)^k)^m (1-(1-\mu)^{k-1})^{N-m}.
\end{split}
\end{equation*}

Using the formula with $x=(1-(1-\mu)^k)(1-\mu/3)$ and
$y = (1-(1-\mu)^{k-1})\mu/3$, we obtain
\begin{equation*}
{N \choose n} (1-\mu)^{kn} (1-\mu/3)^n
\bigg( \big[ x+y \big]^{N-n}
-\big[ (1-(1-\mu)^k)(1-\mu/3)\big]^{N-n} \bigg).
\end{equation*}

Computing the difference between the terms, the probability we are
looking for comes out as
\begin{equation*}
{N \choose n} (1-\mu)^{kn} (1-\mu/3)^n \bigg(
\big[ 1-(1-\mu)^k \big]^{N-n} -
\big[1-(1-\mu+\mu^2/3)(1-\mu)^{k-1} \big]^{N-n} \bigg).
\end{equation*}

One can verify that the sum of the terms above for $n=1, 2, \ldots, N-1$
is equal to $E_k$.

To find the probabilities of interest, it is useful to precompute the
terms
\begin{equation*}
\begin{split}
E_{k,m} = \sum_{n-1}^{N-1}{N \choose n} (1-\mu)^{kn} (1-\mu/3)^n \times \\
\bigg( \big[ 1-(1-\mu)^k \big]^{N-n} -
\big[1-(1-\mu+\mu^2/3)(1-\mu)^{k-1} \big]^{N-n} \bigg) \times
(1-(1-\mu)^m)^n.
\end{split}
\end{equation*}

Using Newton's binomial formula we find
\begin{equation*}
\begin{split}
E_{k,m} = \big[ (1-\mu)^k(1-\mu/3)(1-(1-\mu)^m) + 1-(1-\mu)^k \big]^N
- \big[1-(1-\mu)^k \big]^N + \\
\big[ 1-(1-\mu+\mu^2/3)(1-\mu)^{k-1} \big]^N - \\
\big[(1-\mu)^k(1-\mu/3)(1-(1-\mu)^m) +
1-(1-\mu+\mu^2/3)(1-\mu)^{k-1} \big]^N.
\end{split}
\end{equation*}

Note that $E_{k,0} = 0$.

The probability that there is a $\eqg$ symbol among the first $k$
nucleotides after a $\dn$ symbol is $q^k H_k / (1-(1-\mu/3)^N)$, where
\begin{eqnarray*}
H_k &=&
\sum_{m=1}^{N-1}{N \choose m}
(1-\mu/3)^m(\mu/3)^{N-m} \big(1-(1-\mu)^k\big)^{N-m}
\big(1 - \big( 1-(1-\mu)^k \big)^m\big) \\
&=& [1-\mu/3(1-\mu)^k]^N - [1-\mu/3]^N - [1-(1-\mu)^k]^N +
[(1-\mu/3)(1-(1-\mu)^k)]^N.
\end{eqnarray*}

The probability that the $k$-th symbol after a $\dn$ symbol is also $\dn$
is
\begin{equation*}
\big( 1-q^kH_k/(1-(1-\mu/3)^N) \big) \cdot p(1-(1-\mu/3)^N) =
p(1-(1-\mu/3)^N) - pq^kH_k.
\end{equation*}

The probability that the $k$-th symbol is $\DN$ is
\begin{equation*}
\big( 1-q^kH_k/(1-(1-\mu/3)^N) \big) \cdot p(1-\mu/3)^N =
p(1-\mu/3)^N - pq^kH_k \frac{(1-\mu/3)^N}{1-(1-\mu/3)^N}.
\end{equation*}
